Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-03-13 WholePlotTrajectories.Rmd
Latests update before merge

@@

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas other less common species were identified under two identifiers only separating trees and palm trees. 
The botanical campaigns carried every 5 to 6 years to identify all trees at the species level only started in 2003 and identification practices varied among plots and successive campaigns. 
This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that had to be propagated to composition and diversity metrics throught a Bayesian framework. 
The uncertainty propagation was done by the replenishment of inventories completed at genus level from real incomplete ones on the basisof vernacular/botanical names association. 
Vernacular names were replaced through multinomial trials (M~v~([*s~1~, s~2~, …, s~N~*] ,[*α~1~, α~2~,…, α~N~*]) ) based on the observed association probability [*α~1~, α~2~,…, α~N~*] between each vernacular name *v* and the species [*s~1~, s~2~, …, s~N~*] recorded in the inventory. 
To avoid remaining identification caveats and consider complete inventories, the simulated inventories were then reported at genus.
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology.

The functional diversity metrics used a dataset for 6 functional traits representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area, the leaf area per unit dry mass) and wood economics (wood specific gravity and bark thickness), and life history traits (maximum specific height and seed mass).
Traits database came from the BRIDGE project (http://www.ecofog.gf/Bridge/index.html) where traits values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou. 
Missing trait values were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of the functional traits. 
The dataset comprised 294 botanical species pertaining to 157 botanical genus. 
Whenever a species inventoried was not in the dataset, it was attributed a set of traits values randomly sampled among species of the same next higher taxonomic level. 
As seed mass information correspond to a classification into mass classes, no data filling process was applied so analysis were performed considering the 414 botanical species of the seed mass dataset.
All composition and diversity metrics corresponded to the expectancy obtained after 50 iterations of the taxonomic uncertainty propagation framework and of the filling process of missing trait values.

@@

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973]. 
To tackle the unequal number of recruited trees among treatments the indices bias corrected estimator were used, following [@chao2015estimating;@Marcon2015b].
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), which proved well suited for diversity studies [@Patil1982;@Tothmeresz1995]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and life traits.

@@

*The distance from the initial condition was significantly time-dependent (P<0.01).* 
*The coordinates of the functional traits were measured in the two-dimensional ordination space that mapped plots evolution along time (graphs not showed) which revealed that the trajectories of disturbed plots along time headed toward acquisitive functional strategies (from high SWG to high SLA and chlorophyll content).*

@@

In line with the assumtions of the IDH the observed diversity trajectories confirmed the increase of communities evenness and functional richness after disturbance.
The consistency of the IDH has been debated for the higly diverse tropical forests, some observing the importance of environmental filters to shape communities assembly and others arguing for a genuine stochastic recovery [@Hubbell1999;@Molino2001].
Consistently to the trajectories drawn here, a previous study of the same experiment revealed the directional changes in communities composition reflecting environmental filtering after disturbance [@Molino2001].
Disturbance regimes brought a unimodal response depending on disturbance intensity as suggested by the outcome of 10 years and 20 years post-disturbance analysis[@Molino2001;@Baraloto2012a].
As observed after logging in a Boranean rainforest [@Cannon1998] or after cyclones in the Solomon islands [@Burslem2000], communities diversity showed a unimodal response to disturbance and a maintenance of diversity levels, except for one of the intensively disturbed community which evenness was still increasing \@ref(fig:Fig5). 
Despite the recovery of communities diversity, though, taxonomic composition was clearly altered \@ref(fig:Fig3) implying a long-term, if not permenant, composition change after disturbance as observed in East African tropical forests [@Bonnell2011].

@@

The diversity and composition functional trajectories showed a decoupling between functional and taxonomic after the most intense disturbance (treatments T2 and T3).
Plots functional characteristic returned to their initial values, similar to those of control plots, faster than their taxonomic counterparts \@ref(fig:Fig2;fig:Fig3;fig:Fig4).
As functional traits are the most direct link between biodiversity and ecosystem functioning [@Diaz2005] the recovery of communities functional characteristics would go along the recovery of ecosystems processes although taxonomic composition remain altered [@Guariguata2001].
Such decoupling suggested the consistency of functional overlap and maybe functional versatility, implying a variety of functioning from the same characteristic, in the pre-disturbance condition.
Functional overlap has been observed in highly diverse ecosystems [@Bellwood2006] and suggested as one mechanisms for high diversity maintenance.
Knowing the decoupling between taxonomic and functional trajectories, this functional overlap may not be recovering after disturbance and therefore radically modify communities resilience and resistance [@Trenbath1999;@Elmqvist2003;@Diaz2005]. 
Given the impaired functional overlap, the persitant compositional changes regarding disturbance resistant species [@Haddad2008] and probably lianas or epiphytes [@Martin2013], and the modification of environmental conditions like soils nutrient cycling and compaction [@Olander2005] the entirety of communities recovery after intense disturbance is deeply questioned [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Although communities recovered the core of their initial functioning, composition seemed permanently altered after intense disturbance.
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Diaz2005;@Gardner2007].

>>>>

```{r kable}
textExp<- c(" ","Timber", "Thinning", "Fuelwood", "\\%AGB lost", 
         "T1","DBH ≥ 50 cm,commercial species,~ 10 trees/ha"," "," ","[12\\%-33\\%]",
         "T2","DBH ≥ 50 cm,commercial species,~ 10 trees/ha","DBH ≥ 40 cm,non-valuable species,~ 30 trees/ha",
            " ","[33\\%-56\\%]",
         "T3","DBH ≥ 50 cm,commercial species,~ 10 trees/ha","DBH ≥ 50 cm,non-valuable species,~ 15 trees/ha",
         "40 cm ≤ DBH ≤ 50 cm,non-valuable species,~ 15 trees/ha","[35\\%-56\\%]",)
knitr::kable(textExp, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou", booktabs = TRUE)
```

@@

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas other less common species were identified under two identifiers only separating trees and palm trees. 
The botanical campaigns carried every 5 to 6 years to identify all trees at the species level only started in 2003 and identification practices varied among plots and successive campaigns. 
This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that had to be propagated to composition and diversity metrics throught a Bayesian framework. 
The uncertainty propagation was done by the replenishment of inventories completed at genus level from real incomplete ones on the basisof vernacular/botanical names association. 
Vernacular names were replaced through multinomial trials (M~v~([*s~1~, s~2~, …, s~N~*] ,[*\alpha~1~, \alpha~2~,…, \alpha~N~*]) ) based on the observed association probability [*\alpha~1~, \alpha~2~,…, \alpha~N~*] between each vernacular name *v* and the species [*s~1~, s~2~, …, s~N~*] recorded in the inventory. 
To avoid remaining identification caveats and consider complete inventories, the simulated inventories were then reported at genus.
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology.

The functional diversity metrics used a dataset for 6 functional traits representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area, the leaf area per unit dry mass) and wood economics (wood specific gravity and bark thickness), and life history traits (maximum specific height and seed mass).
Traits database came from the BRIDGE project [^1] where traits values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou. 
Missing trait values were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of functional traits. 
The dataset comprised 294 botanical species pertaining to 157 botanical genus. 
Whenever a species inventoried was not in the dataset, it was attributed a set of traits values randomly sampled among species of the same next higher taxonomic level. 
As seed mass information correspond to a classification into mass classes, no data filling process was applied so analysis were performed considering the 414 botanical species of the seed mass dataset.
All composition and diversity metrics corresponded to the expectancy obtained after 50 iterations of the taxonomic uncertainty propagation framework and of the filling process of missing trait values.

[^1] http://www.ecofog.gf/Bridge/

@@

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973]. 
To tackle the unequal number of recruited trees among treatments the indices bias corrected estimator were used, following [@Chao2015;@Marcon2015b].
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), which proved well suited for diversity studies [@Patil1982;@Tothmeresz1995]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and life traits.

@@

>The distance from the initial condition was significantly time-dependent (P<0.01).
>The coordinates of the functional traits were measured in the two-dimensional ordination space that mapped plots evolution along time (graphs not showed) which revealed that the trajectories of disturbed plots along time headed toward acquisitive functional strategies (from high SWG to high SLA and chlorophyll content).

@@

In line with the assumtions of the IDH the observed diversity trajectories confirmed the increase of communities evenness and functional richness after disturbance.
The consistency of the IDH has been debated for the higly diverse tropical forests, some observing the importance of environmental filters to shape communities assembly and others arguing for a genuine stochastic recovery.
Consistently to the trajectories drawn here, a previous study of the same experiment revealed the directional changes in communities composition reflecting environmental filtering after disturbance.
Disturbance regimes brought a unimodal response depending on disturbance intensity as suggested by the outcome of 10 years and 20 years post-disturbance analysis[@Hubbell1999;@Molino2001;@Baraloto2012a].
As observed after logging in a Boranean rainforest [@Cannon1998] or after cyclones in the Solomon islands [@Burslem2000], communities diversity showed a unimodal response to disturbance and a maintenance of diversity levels, except for one of the intensively disturbed community which evenness was still increasing \@ref(fig:Fig5). 
Despite the recovery of communities diversity, though, taxonomic composition was clearly altered \@ref(fig:Fig3) implying a long-term, if not permenant, composition change after disturbance as observed in East African tropical forests [@Bonnell2011].

@@

The diversity and composition functional trajectories showed a decoupling between functional and taxonomic after the most intense disturbance (treatments T2 and T3).
Plots functional characteristic returned to their initial values, similar to those of control plots, faster than their taxonomic counterparts \@ref(fig:Fig2;fig:Fig3;fig:Fig4).
As functional traits are the most direct link between biodiversity and ecosystem functioning [@Diaz2005] the recovery of communities functional characteristics would go along the recovery of ecosystems processes although taxonomic composition remain altered [@Guariguata2001].
Such decoupling suggested the consistency of functional overlap and maybe functional versatility, implying a variety of functioning from the same characteristic, in the pre-disturbance condition.
Functional overlap has been observed in highly diverse ecosystems [@Bellwood2006] and suggested as one mechanisms for high diversity maintenance.
Knowing the decoupling between taxonomic and functional trajectories, this functional overlap may not be recovering after disturbance and therefore radically modify communities resilience and resistance [@Trenbath1999;@Elmqvist2003;@Diaz2005]. 
Given the impaired functional overlap, the persitant compositional changes regarding disturbance resistant species [@Haddad2008] and probably lianas or epiphytes [@Martin2013], and the modification of environmental conditions like soils nutrient cycling and compaction [@Olander2005] the entirety of communities recovery after intense disturbance is deeply questioned [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Although communities recovered the core of their initial functioning, composition seemed permanently altered after intense disturbance.
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].
